Perceived value of video games, but not hours played, predicts mental well-being in causal adult Nintendo players

### 2024-12-02 ms.qmd
revise for R1

---
title: "Perceived value of video games, but not hours played, predicts mental well-being in adult Nintendo players"
shorttitle: "Nintendo gaming and well-being"
author:
  - name: Nick Ballou
    corresponding: false
    orcid: 0000-0003-4126-0696
    email: nick@nickballou.com
    affiliations:
      - name: University of Oxford
        department: Oxford Internet Institute
    # roles:
    #   - conceptualization
    #   - data curation
    #   - methodology
    #   - formal analysis
    #   - writing

@@

  - name: Andrew K Przybylski
    corresponding: true
    email: andy.pryzbylski@oii.ox.ac.uk
    orcid: 0000-0003-4126-0696
    affiliations:
      - name: University of Oxford
        department: Oxford Internet Institute
    # roles:
    #   - conceptualization
    #   - funding acquisition
    #   - project administration
    #   - editing
abstract: "Studies on video games and well-being often rely on self-report measures or data from a single game. Here, we study how 703 US adults' time spent playing over 150 1st-party Nintendo Switch games (totaling 140k hours) relates to their life satisfaction, affect, depressive symptoms, and general mental well-being. We replicate previous findings that playtime over the past 2 weeks does not predict well-being, and extend these findings to a wider range of timescales (1 hour to 1 year). Results suggest that relationships, if present, dissipate within 2 hours of gameplay. Our non-causal findings suggest substantial confounding would be needed to shift a meaningful true effect to the observed null. Although playtime was not related to well-being, players’ assessments of the value of game time—so called gaming life fit—was. Results emphasize the importance of defining the gaming population of interest, collecting data from more than one game, and focusing on how players integrate gaming into their lives rather than the amount of time spent."
keywords: [video games, digital trace data, well-being]
author-note:
  # Disclosures condensed to one paragraph, but you can start a field with two line breaks to break them up: \n\nNew Paragraph
  disclosures:
    study-registration: This study was registered on the Open Science Framework (https://osf.io/sjqyt).
    # Acknowledge and cite data/materials to be shared.
    data-sharing: Data and materials are available at https://osf.io/6xkdg/. 
    # Example: [Author name] has been a paid consultant for Corporation X, which funded this study.
    conflict-of-interest: Data for this study was provided by Nintendo of America. Nintendo of America had no role in the design of the study, the analysis of the data, or decision to publish any results. The authors declare no other competing interests. 
    # Example: This study was supported by Grant [Grant Number] from [Funding Source].
    financial-support: This research was supported by Huo Family Foundation and the UK Economic and Social Research Council (ES/W012626/1). KM was supported by Forskningsrå det för hälsa, arbetsliv och välfärd (2021-01284).
    # Example: The authors are grateful to [Person] for [Reason].
    gratitude: ~
    # Example. Because the authors are equal contributors, order of authorship was determined by a fair coin toss.
    authorship-agreements: ~
bibliography: references.bib
floatsintext: true # If true, tables and figures are mingled with the text instead of listed at the end of the document.
numbered-lines: false # Numbered lines (.pdf and .docx only)
meta-analysis: false
suppress-credit-statement: true
format:
  # apaquarto-docx: default
    # blank-lines-above-author-note: 0
    # blank-lines-above-title: 1
  # apaquarto-html:
  #   documentmode: man
  #   code-fold: true
  #   code-summary: "Show the code"
  #   wordcount-html: default
  apaquarto-pdf:
    documentmode: jou
---

@@

```{r}
#| label: load-libraries

library(tidyverse)
library(lubridate)
library(knitr)
library(kableExtra)
library(dtplyr)
library(scales)
library(ggdist)
library(distributional)
library(mgcv)
library(patchwork)
library(see)
library(performance)
library(careless)
library(cowplot)
library(marginaleffects)
library(flextable)
library(ggstance)
library(crosstable)

@@

Video games played on smartphones, computers, or home consoles are now a widely pursued form of leisure that involves social interaction, creativity, problem-solving, and growth [@BourgonjonEtAl2016Players]. Major firms like Nintendo have sold hundreds of millions of games consoles in recent years [@NintendoJapan2024Top] and online platforms like Steam typically have more than 30 million players online at any given time [@SteamCharts2024SteamCharts]. This staggering investment of human capital has inspired both national [@AmericanPsychiatricAssociation2013diagnostic] and international [@WorldHealthOrganization2018international] health bodies to focus on play as contributor to psychopathology.

@@

In matters of health policy, overall time spent with games—regardless of what or why one plays—remains central to how games are thought to influence player outcomes. Parental control tools on platforms like Xbox and PlayStation foreground time limits as a primary means of enforcing healthy gaming behavior [@Robertson2021Taming]; for adults, a growing array of self-control tools, apps, and dashboards offer the ability to "limit or cut yourself off from apps and games" ([https://focusme.com](https://focusme.com/how-it-works/)), savings users "1.23 hours" (<https://www.opal.so>) or "up to 2.5 hours a day" ([https://freedom.to/](https://freedom.to/freedom-for-ios)). News media suggest time-based limits [e.g., @Saveva2023Healthy], often referencing the since-abandoned 2x2 rule from the American Academy of Pediatrics: no screen time for children under 2, and no more than 2 hours per day for children older than 2 [@Blum-RossLivingstone2018Trouble]. Likewise, the American Psychiatric Association’s description of Internet Gaming Disorder characterizes pathological engagement with games, in part, as involving ‘8-10 hours or more per day \[and or\] least 30 hours per week’ [@AmericanPsychiatricAssociation2013diagnostic, p. 796]. On a larger scale, time-based limits such as Korea’s 10-year Youth Protection Revision Act prohibited young people from playing games between 00:00 and 06:00 [@SangEtAl2017Mobile]. More recently, China put in place a three-hour weekly limit for under-18s [@FeinerKharpal2021China]. The effectiveness of such regulatory steps has been challenged [@zendle2023; @ChoiEtAl2018Effect].

@@

Only a handful of studies have applied digital trace data to the study of games and well-being \[ @VuorreEtAl2022Time; @JohannesEtAl2021video; @BruhlmannEtAl2020Motivational; @LarrieuEtAl2023How\], in part because this data can be very difficult to acquire: researchers must build or rely on (often unstable) technical systems to log data themselves, or negotiate individual agreements with games companies who have historically been reluctant to share data [@Ballou2023Manifesto; @SeifEl-NasrEtAl2013game]. Where digital trace data has been collected, however, results have been informative. @BruhlmannEtAl2020Motivational used playtime and in-game behavior to identify subgroups of *League of Legends* players who had more negative in-game experiences. @JohannesEtAl2021video look at playtime in *Animal Crossing: New Horizons* and *PvZ: Battle for Neighborville* and found a positive but likely negligible correlation. A follow-up study expanded this to 7 games, finding that changes in playtime over the course of 6 weeks were unlikely to affect subsequent well-being [@VuorreEtAl2022Time]. @LarrieuEtAl2023How follow high-intensity *Rainbow 6: Siege* players and find no link between playtime and quality of life across any identified player types.

Though informative, this early work has a vital scope limitation: digital trace data was only available for a single game per player. It was not possible to know what other games a participant was, or was not, electing to play. Players regularly switch between games over the course of a day or week based on mood, available time, and social context [@ONeillEtAl2016Condensing; @BallouEtAl2024Registered]; data collected for one particular game may thus tell us little about *overall* playtime or its relation to well-being.

An important frontier for the field, therefore, is to collect holistic digital trace data that reflects behavior in not just one game, but all games played (which may include various games on one platform, such as Nintendo Switch or Steam, or all games played across several platforms a player uses). At present, our understanding of even basic phenomena such as the true volume of play in different demographics rely on the same inaccurate self-report data, itself often provided by market research firms using opaque methodologies. Capturing play at the platform level represents one step towards this goal of studying a player's complete gaming diet. To our knowledge, the only study to platform-level digital trace data to investigate player health is [@BallouEtAl2024Registered], which found no meaningful relationships between Xbox playtime and well-being over 3 months.

@@

```{r}
#| label: fig-sankey
#| fig-align: center
#| out-width: "80%"
#| fig-cap: "Participant flow, from recruitment to final sample"
#| apa-twocolumn: true

knitr::include_graphics("figures/sankey.png")
```

In the present research we report on a study conducted in collaboration with Nintendo of America, in which we independently recruited a large sample of adult video game players, surveyed them about their motivation and well-being, and joined these responses to digital trace data on Nintendo Switch video game play. Our central aim was to test the extent to which the amount of time participants spent playing related to their psychological well-being. Our analysis plan was preregistered at <https://osf.io/sjqyt>.

@@

We were also interested in understanding how relationships between play and well-being might depend on the choice of what time scale of play is considered relevant. To this end, we examined relationships between well-being and a wide range of time windows of play preceding the well-being question. While preregistered, the large number of models implicated in our analysis plan is prohibitively large to be aptly framed as a narrow hypothesis test. We therefore operationalize this as an exploratory research question: How does the relationship between playtime and well-being differ across different gaming observation windows ranging from 1 hour to 1 year?

Lastly, we investigated what factors might moderate the relationship between playtime and well-being. To test this exploratory question, we assessed potential moderation by gender, age, as well as a subjective sense of how players thought gaming had positive or negative relationships to various life domains such as work, relationships, school performance, and social health.

@@

# filter out in stages, in case of manual look at change in sample size after each
dat <- datAll |>
  filter(!pid %in% nonResponders) |>
  filter(!pid %in% nonPlayers) |>
  filter(!pid %in% clockHackers) |>
  filter(possibleCareless %in% c(FALSE, NA)) |>
  select(-starts_with(c("wemwbs", "bangs", "trojan", "promis", "timeUse", "recent", "weird")), -c(positives, problematicPlay, feedback))

@@

Our participant flow is shown in @fig-sankey. We recruited participants from Prolific, a participant recruitment platform, who were: (1) 18 years old, (2) proficient in English, (3) residents of the US, and (4) active video game players (self-defined, based on a \>0 response to Prolific's built-in screening item "How many hours per week do you play video games on average?"). We first distributed a screening questionnaire to 7649 participants asking which video game platforms they were active on; of these, 4184 indicated that they played games on Nintendo Switch.

We invited these participants to a second survey wherein they retrieved an account identifier from the Nintendo web interface (the events QR code, available at <https://accounts.nintendo.com/qrcode>), and shared it with us. This identifier is separate from their username, and cannot be used by anyone besides Nintendo (including ourselves) to link the player to personally identifiable information. A total of 1823 participants completed the linking process. We sent the account identifiers to Nintendo of America, who in turn sent us each player's pseudonymized play history from May 1, 2022 to present. Data collection began with a pilot study of 100 participants on Nov 15, 2023, which was combined with primary data collection from Feb 12, 2024 to May 6, 2024.

Of the participants who completed the linking process, 1607 had eligible Nintendo data (i.e., any play sessions of a 1st party game—a game published by Nintendo or one of its subsidiaries, as opposed to a third-party publisher—from May 1, 2022 to present). These 1607 were subsequently sent a well-being survey in Qualtrics (see Measures below). 1191 participants completed the well-being survey; our preregistered stopping rule went into effect when 5 or fewer participants per day completed the survey for 3 consecutive days.

As preregistered, we excluded 427 individuals who had 0 minutes of playtime during the previous 3 months, indicating that they are not active Nintendo players, and 26 people who logged more than 24 hours of playtime on any single day or who had a session that took place in the future, indicating a technical problem or manipulation of the system clock for in-game benefits. We further excluded 34 participants who displayed signs of careless responding (e.g., so-called straightlining or seemingly random responses). In total, we excluded 487 participants (some participants were excluded on multiple grounds), leading to a final sample of `r nrow(dat)`.

Participants were paid £0.15 for the 1-minute screening questionnaire, £3 for linking their data (\~5 minutes) plus a £2 bonus payment if data was successfully retrieved, and £4 for a 22-minute well-being survey. This study received approval from the University of Oxford Social Sciences and Humanities Interdivisional Research Ethics Committee (OII_C1A_23_107).

@@

Participants completed a self-report survey that took on average 22 minutes to complete. The survey included background factors such as demographics and life circumstances, a series of cognitive tasks, as well as self-report measures of time use, and motivations for video game play. We detail those measures we used in this study below but all study data are available at OSF (see supplementary materials).

**Video game playtime** was measured by collecting a record of each player's individual game sessions for all 1st party video games played on a Nintendo Switch. These data were provided by Nintendo of America. Playtime was calculated by summing the duration of all (partial) sessions that occur during a given time period relative to the participant's survey response, based on the logged session start and end times. For ease of interpretation, game play time in all observation periods longer 24 hours is reported as mean minutes of play per day. It is important to note that these data included only telemetry for games published by Nintendo and its close business partners (e.g. The Pokémon Company), but not games made by third party publishers (e.g. Electronic Arts). Nintendo-published games accounted for 63% of all playtime across our sample; the 37% of play data from 3rd-party games is therefore treated as missing. We return to this limitation in the discussion.

@@

**Life satisfaction** was measured with the one-item Cantril self-anchoring scale [@Cantril1965pattern]. Participants were prompted with ‘Please imagine a ladder with steps numbered from 0 at the bottom to 10 at the top. The top of the ladder represents the best possible life for you, and the bottom of the ladder represents the worst possible life for you. On which step of the ladder would you say you personally feel you stood over the past two weeks?’. Participants responded on a scale from 0 to 10, which was rescaled to 1-5 to match the other well-being measures.

**Affect** was measured with a single item: "How are you feeling right now?" [@KillingsworthGilbert2010Wandering]. Participants responded using a 100-point visual analogue scale (VAS) with endpoints "Very bad" and "Very good", which we rescaled to 1-5 to match the other well-being measures.

@@

To test H1, we fit multiple regression models with playtime over the previous 2 weeks as the primary predictor, all demographic variables as covariates (age, gender, highest level of education, and employment status), and the corresponding mental health variable as the outcome. For example, for H1a (life satisfaction predicted by the previous 2 weeks of play), the model in R is:

@@

We apply a similar analysis approach to our exploration of H2 concerning other timescales; we primarily apply multiple regression with well-being predicted by playtime aggregated over various time periods and the same covariates, but also explore potential non-linear alternatives and moderation analyses (detailed below).

We interpret the playtime coefficient estimates from these models in reference to pre-specified smallest effect sizes of interest (see below): if the 99% confidence interval is fully within the upper and lower equivalence bounds, this provides evidence to reject a practically meaningful association.

We conducted all statistical analyses with R version 4.3.2 [@RCoreTeam2023language]. We use an alpha of .01.

@@

Note: this method of specifying an SESOI is predicated on a causal interpretation—it implicitly imagines a world where one can intervene on the predictor (playtime) and have an effect of a certain size on the outcome (mental health). It is very unlikely that our cross-sectional analyses can provide unbiased causal estimates. Instead, our goal is to use associations to place boundaries on the size of a possible effect: if there is no meaningful correlation between playtime and mental health, there is even less likely to be a meaningful causal effect between playtime and mental health. We support this reasoning with simulations presented in the discussion.

@@

```{r}
#| label: fig-playtime
#| fig-cap: "Description of playtime in the sample. Panel A shows the distribution of playtime across players for each observation period, with the mean shown in red; observations periods below the dashed line are shown as total hours, whereas periods above the dashed line reflect hours of player per day. Panel B shows the proportion of players who logged at least one play session in that period."
#| fig-width: 8
#| fig-height: 4
#| fig-align: center
#| fig-pos: "t"
#| apa-twocolumn: true

@@

Given the lack of holistic or platform-level data available in the literature, our first goal was to simply describe the volume of play. This is visualized in @fig-playtime, which show that despite a total play volume of more than 140k hours, our sample was largely minimally engaged with 1st party Nintendo games. During the 2 weeks prior to survey completion, just over half of the sample had 0 sessions logged. The top 10% of players were moderately engaged, playing an average of 60 minutes per day. Sessions of a game lasted on average `r round(median(datRaw$duration, na.rm = T), 1)` minutes \[10th percentile: `r quantile(datRaw$duration, .1)`; 90th percentile: `r quantile(datRaw$duration, .9)`\].

The results of this study are therefore reflective of a largely casual population of players—at least with respect to Nintendo titles. We argue that this population is nonetheless an important one: if video games were to meaningfully affect well-being, we may expect a larger impact for people who rarely play but happen to play for 1 hour, than for a highly engaged population of people who tend to play 3 hours per day, but happen to play 4 hours. We return to this limitation on generalizability in the discussion.

@@

```{r}
#| label: model-fitting

@@

Next, we conducted exploratory analyses to investigate what other factors might influence who exhibits positive or negative relations between gaming and well-being. We explored age, gender, and life fit---the perceived harmful or beneficial value of gaming in various life domains. We expected that people who perceive gaming to support other life domains would exhibit a more positive relationship between gaming and mental health, and those who perceived gaming to be harmful to other life domains would exhibit a more negative relationship.

>>>>

---
title: "Perceived value of video games, but not hours played, predicts mental well-being in causal adult Nintendo players"
shorttitle: "Nintendo gaming and well-being"
author:
  - name: Nick Ballou
    corresponding: false
    orcid: 0000-0003-4126-0696
    email: nick@nickballou.com
    affiliations:
      - name: University of Oxford
        department: Oxford Internet Institute
    # roles:
    #   - conceptualization
    #   - data curation
    #   - methodology
    #   - formal analysis
    #   - writing

@@

  - name: Andrew K Przybylski
    corresponding: true
    email: andy.pryzbylski@oii.ox.ac.uk
    orcid: 0000-0003-4126-0696
    affiliations:
      - name: University of Oxford
        department: Oxford Internet Institute
    # roles:
    #   - conceptualization
    #   - funding acquisition
    #   - project administration
    #   - editing
abstract: "Studies on video games and well-being often rely on self-report measures or data from a single game. Here, we study how 703 casually-engaged US adults’ time spent playing for over 140,000 hours across 150 Nintendo Switch games relates to their life satisfaction, affect, depressive symptoms, and general mental well-being. We replicate previous findings that playtime over the past two weeks does not predict well-being, and extend these findings to a wider range of timescales (one hour to one year). Equivalence tests were inconclusive, and thus we do not find evidence of absence, but results suggest that relationships, if present and practically meaningful, likely dissipate within two hours of gameplay. Our non-causal findings suggest substantial confounding would be needed to shift a meaningful true effect to the observed null. Although playtime was not related to well-being, players’ assessments of the value of game time—so called gaming life fit—was. Results emphasise the importance of defining the gaming population of interest, collecting data from more than one game, and focusing on how players integrate gaming into their lives rather than the amount of time spent."
keywords: [video games, digital trace data, well-being]
author-note:
  # Disclosures condensed to one paragraph, but you can start a field with two line breaks to break them up: \n\nNew Paragraph
  disclosures:
    study-registration: This study was registered on the Open Science Framework (https://osf.io/sjqyt).
    # Acknowledge and cite data/materials to be shared.
    data-sharing: Data and materials are available at https://osf.io/6xkdg/. 
    # Example: [Author name] has been a paid consultant for Corporation X, which funded this study.
    conflict-of-interest: Data for this study was provided by Nintendo of America. Nintendo of America had no role in the design of the study, the analysis of the data, or decision to publish any results. The authors declare no other competing interests. 
    # Example: This study was supported by Grant [Grant Number] from [Funding Source].
    financial-support: This research was supported by Huo Family Foundation and the UK Economic and Social Research Council (ES/W012626/1). KM was supported by Forskningsrå det för hälsa, arbetsliv och välfärd (2021-01284).
    # Example: The authors are grateful to [Person] for [Reason].
    gratitude: ~
    # Example. Because the authors are equal contributors, order of authorship was determined by a fair coin toss.
    authorship-agreements: ~
bibliography: references.bib
floatsintext: true # If true, tables and figures are mingled with the text instead of listed at the end of the document.
numbered-lines: true # Numbered lines (.pdf and .docx only)
meta-analysis: false
suppress-credit-statement: true
format:
  apaquarto-docx: default
  # apaquarto-html:
  #   documentmode: man
  #   code-fold: true
  #   code-summary: "Show the code"
  #   wordcount-html: default
  apaquarto-pdf:
    documentmode: jou
---

@@

```{r}
#| label: load-libraries
if (!require("pacman")) install.packages("pacman")
library(pacman)

p_load(tidyverse, lubridate, knitr, kableExtra, dtplyr, scales, ggdist, distributional, mgcv,
       patchwork, see, performance, careless, cowplot, marginaleffects, flextable, ggstance, crosstable,
       lavaan, glue)

@@

Video games played on smartphones, computers, or home consoles are now a widely pursued form of leisure that involves social interaction, creativity, problem-solving, and growth [@BourgonjonEtAl2016Players]. Major firms like Nintendo have sold hundreds of millions of games consoles in recent years [@NintendoJapan2024Top] and online platforms such as Steam regularly attract upwards of 30 million players online at any given time [@SteamCharts2024SteamCharts]. This staggering investment of human attention and behaviour has inspired both national [@AmericanPsychiatricAssociation2013diagnostic] and international [@WorldHealthOrganization2018international] health bodies to focus on play as potential contributor to psychopathology.

@@

In matters of health policy, overall time spent with games—regardless of what or why one plays—remains central to how games are thought to influence player outcomes. Parental control tools on platforms like Xbox and PlayStation foreground time limits as a primary means of enforcing healthy gaming behavior [@Robertson2021Taming]; for adults, a growing array of self-control tools, apps, and dashboards offer the ability to "limit or cut yourself off from apps and games" ([https://focusme.com](https://focusme.com/how-it-works/)), savings users "1.23 hours" (<https://www.opal.so>) or "up to 2.5 hours a day" ([https://freedom.to/](https://freedom.to/freedom-for-ios)). News media suggest time-based limits [e.g., @Saveva2023Healthy], often referencing the since-abandoned 2x2 rule from the American Academy of Pediatrics: no screen time for children under two, and no more than two hours per day for children older than two [@Blum-RossLivingstone2018Trouble]. Likewise, the American Psychiatric Association’s description of Internet Gaming Disorder characterises pathological engagement with games, in part, as involving ‘8-10 hours or more per day \[and or\] least 30 hours per week’ [@AmericanPsychiatricAssociation2013diagnostic, p. 796]. On a larger scale, time-based limits such as Korea’s 10-year Youth Protection Revision Act prohibited young people from playing games between 00:00 and 06:00 [@SangEtAl2017Mobile]. More recently, China put in place a weekly three-hour limit for under-18s [@FeinerKharpal2021China]. The effectiveness of such regulatory steps has been challenged [@zendle2023; @ChoiEtAl2018Effect].

@@

Only a handful of studies have applied digital trace data to the study of games and well-being [see @VuorreEtAl2022Time; @JohannesEtAl2021video; @BruhlmannEtAl2020Motivational; @LarrieuEtAl2023How], in part because this data can be very difficult to acquire: researchers must build or rely on unstable technical systems to log data themselves, or negotiate individual agreements with games companies who have historically been reluctant to share data [@Ballou2023Manifesto; @SeifEl-NasrEtAl2013game]. Where digital trace data has been collected, however, results have been informative. @BruhlmannEtAl2020Motivational used playtime and in-game behavior to identify subgroups of *League of Legends* players who had more negative in-game experiences. @JohannesEtAl2021video look at playtime in *Animal Crossing: New Horizons* and *PvZ: Battle for Neighborville* and found a positive but likely negligible correlation. A follow-up study expanded this to seven games, finding that changes in playtime over the course of 6 weeks were unlikely to affect subsequent well-being [@VuorreEtAl2022Time]. @LarrieuEtAl2023How follow high-intensity *Rainbow 6: Siege* players and find no link between playtime and quality of life across any identified player types.

Though informative, this early work has a vital scope limitation: digital trace data was only available for a single game for each player-participant. It was not possible to know what other games a participant was, or was not, electing to play. Players regularly switch between games over the course of a day or week based on mood, available time, and social context [@ONeillEtAl2016Condensing; @BallouEtAl2024Registered]; data collected for one particular game may thus tell us little about *overall* playtime or its relation to well-being.

An important frontier for the field, therefore, is to collect holistic digital trace data that reflects behavior in not just one game, but all games played (which may include various games on one platform, such as Nintendo Switch or Steam, or all games played across several platforms a player uses). At present, our understanding of even basic phenomena such as the true volume of play in different demographics rely on the same inaccurate self-report data, itself often provided by market research firms using opaque methodologies. Capturing play at the platform level represents one step towards this goal of studying a player's complete gaming diet. To our knowledge, the only study to investigate player health with platform-level digital trace data is [@BallouEtAl2024Registered], which found no meaningful relationships between Xbox playtime and well-being over three months. The current study complements this earlier work by investigating how well-being relates to play on an alternative platform, Nintendo Switch, which has a more causal audience and a distinct library of games, and by analyzing a longer window of play behavior spanning up to one year.

@@

```{r}
#| label: fig-sankey
#| fig-align: center
#| out-width: "100%"
#| fig-cap: "Participant flow, from recruitment to final sample"
#| apa-twocolumn: true

knitr::include_graphics("tables-figures/sankey.png")
```

In the present research we report on a study in which we independently recruited a large sample of adult video game players, surveyed them about their motivation and well-being, and joined these responses to digital trace data on Nintendo Switch video game play, which was provided via a data-sharing agreement with Nintendo of America. Our central aim was to test the extent to which the amount of time participants spent playing related to their psychological well-being. Our analysis plan was preregistered at <https://osf.io/sjqyt>.

@@

We were also interested in understanding how relationships between play and well-being might depend on the choice of what timescale of play researchers consider relevant to investigate. To this end, we examined relationships between well-being and a wide range of time windows of play preceding the well-being question. While preregistered, the large number of models implicated in our analysis plan is prohibitively large to be aptly framed as a narrow hypothesis test. We therefore operationalise this as an exploratory research question: How does the relationship between playtime and well-being differ across different gaming observation windows ranging from one hour to one year?

Lastly, we investigated what factors might moderate the relationship between playtime and well-being. To test this exploratory question, we assessed potential moderation by participant gender and age, as well as a subjective sense of how players thought gaming had positive or negative relationships to various life domains such as work, relationships, school performance, and social health.

@@

# filter out in stages, in case of manual look at change in sample size after each
dat <- datAll |>
  filter(!pid %in% nonResponders) |>
  filter(!pid %in% nonPlayers) |>
  filter(!pid %in% clockHackers) |>
  filter(possibleCareless %in% c(FALSE, NA)) |>
  select(-starts_with(c("bangs", "trojan", "timeUse", "recent", "weird")), -c(positives, problematicPlay, feedback))

@@

Our participant flow is shown in @fig-sankey. We recruited participants from Prolific, a participant recruitment platform, who were: (1) at least 18 years old, (2) proficient in English, (3) residents of the US, and (4) active video game players (self-defined, based on a \>0 response to Prolific's built-in screening item "How many hours per week do you play video games on average?"). We first distributed a screening questionnaire to 7649 participants asking which video game platforms they were active on; of these, 4184 indicated that they played games on Nintendo Switch. Participants who played on Nintendo Switch were active on 1.8 additional platforms on average (e.g., also using Playstation and iOS); their Nintendo playtime therefore captures only an (unknown) portion of their overall gaming activity. We return to this limitation in the Discussion.

We invited these participants to a second survey wherein they retrieved an account identifier from the Nintendo web interface using the events QR code, available at <https://accounts.nintendo.com/qrcode>, and they shared these unique identifiers with us. This identifier is separate from their username, and cannot be used by anyone besides Nintendo---including our research team---to link the player to personally identifiable information. A total of 1823 participants completed the linking process. We sent the account identifiers to Nintendo of America, who in turn sent us each player's pseudonymised play history from May 1, 2022 to present. Data collection began with a pilot study of 100 participants on Nov 15, 2023, which was combined with primary data collection from Feb 12, 2024 to May 6, 2024.

Of the participants who completed the linking process, 1607 had eligible Nintendo data (i.e., any play sessions of a 1st party game—a game published by Nintendo or one of its subsidiaries, as opposed to a third-party publisher—from May 1, 2022 to present). These 1607 were subsequently sent a well-being survey in Qualtrics detailed below. 1191 participants completed the well-being survey; our preregistered stopping rule went into effect when 5 or fewer participants per day completed the survey for three consecutive days.

As preregistered, we excluded 427 individuals who had no playtime during the previous three months, indicating that they are not active Nintendo players, and 26 people who logged more than 24 hours of playtime on any single day or who had a session that took place in the future, indicating a technical problem or manipulation of the system clock for in-game benefits. We further excluded 34 participants who displayed signs of careless responding (e.g., so-called straightlining or seemingly random responses). In total, we excluded 487 participants (some participants were excluded on multiple grounds), leading to a final sample of `r nrow(dat)`.

Participants were paid £0.15 for the 1-minute screening questionnaire, £3 for linking their data plus a £2 bonus payment if data was successfully retrieved, and £4 for a 22-minute well-being survey. This study received approval from the University of Oxford Social Sciences and Humanities Interdivisional Research Ethics Committee (OII_C1A_23_107).

@@

Participants completed a self-report survey that took on average 22 minutes to complete. The survey included background factors such as demographics and life circumstances, a series of cognitive tasks, as well as self-report measures of time use, and motivations for video game play. In this paper, we focus on the playtime data and the self-reported well-being measures, and detail only those measures below. Other measures were intended for exploratory analyses beyond the scope of the current paper. All study data and materials are available on the OSF (see supplementary materials).

**Video game playtime** was measured by collecting a record of each player's individual game sessions for all 1st party video games played on a Nintendo Switch. These data were provided by Nintendo of America. Playtime was calculated by summing the duration of all (partial) sessions that occur during a given time period relative to the participant's survey response, based on the logged session start and end times. For ease of interpretation, playtime in all observation periods longer than 24 hours is reported as mean minutes of play per day. It is important to note that these data included only digital trace data for games published by Nintendo and its close business partners (e.g. The Pokémon Company), but not games made by third party publishers (e.g. Electronic Arts). Nintendo-published games accounted for 63% of all playtime across our sample; the 37% of play data from 3rd-party games is therefore treated as missing. We return to this limitation in the discussion.

@@

**Life satisfaction** was measured with the one-item Cantril self-anchoring scale [@Cantril1965pattern]. Participants were prompted with ‘Please imagine a ladder with steps numbered from 0 at the bottom to 10 at the top. The top of the ladder represents the best possible life for you, and the bottom of the ladder represents the worst possible life for you. On which step of the ladder would you say you personally feel you stood over the past two weeks?’. Participants responded on a scale from 0 to 10, which was rescaled to continuous values between 1-5 to match the range of WEMWBS and PROMIS measures.

**Affect** was measured with a single item: "How are you feeling right now?" [@KillingsworthGilbert2010Wandering]. Participants responded using a 100-point visual analogue scale (VAS) with endpoints "Very bad" and "Very good", which we rescaled to continuous values between 1-5 to match the WEMWBS and PROMIS measures.

@@

To test H1, we fit multiple regression models with playtime over the previous 2 weeks as the primary predictor, all demographic variables as covariates (age, gender, highest level of education, and employment status), and the corresponding mental health variable as the outcome. For example, for H1a (life satisfaction predicted by the previous 2 weeks of play), the model in R was:

@@

We applied a similar analysis approach to our exploration of H2 concerning other timescales; we primarily applied multiple regression with well-being predicted by playtime aggregated over various time periods and the same covariates, but also explored potential non-linear alternatives and moderation analyses (detailed below).

We interpreted the playtime coefficient estimates from these models in reference to pre-specified smallest effect sizes of interest (see below): if the 99% confidence interval is fully within the upper and lower equivalence bounds, this provides evidence to reject a practically meaningful association.

We conducted all statistical analyses with R version 4.3.2 [@RCoreTeam2023language]. We used an alpha of .01.

@@

Note: this method of specifying an SESOI is predicated on a causal interpretation—it implicitly imagines a world where one can intervene on playtime (our predictor) and have an effect of a certain size on mental health (our outcomes). It is very unlikely that our cross-sectional analyses can provide unbiased causal estimates. Instead, our goal is to use associations to place boundaries on the size of a possible effect. In other words, if there is no meaningful correlation between playtime and mental health, there is even less likely to be a meaningful causal effect between playtime and mental health. We support this reasoning with simulations presented in the discussion.

@@

```{r}
#| label: fig-playtime
#| fig-cap: "Description of playtime in the sample. Panel A shows the distribution of playtime across players for each observation period, with the mean shown in red; observation periods below the dashed line are shown as total hours, whereas periods above the dashed line reflect hours of player per day. Panel B shows the proportion of players who logged at least one play session in that period. Raw descriptive statistics are available in the supplementary materials (tables-figures/playtimeDescriptives.csv)."
#| fig-width: 8
#| fig-height: 4
#| fig-align: center
#| fig-pos: "t"
#| apa-twocolumn: true

playtimeTable <- dat |> 
  select(matches("playtime\\d")) |> 
  summarise(across(
    everything(),
    list(`Mean hours (per day)` = ~ mean(.x, na.rm = TRUE),
         SD = ~ sd(.x, na.rm = TRUE),
         `Median hours (per day)` = ~ median(.x, na.rm = TRUE))
  )) |> 
  pivot_longer(cols = everything(),
               names_to = c("Variable", ".value"),
               names_sep = "_") |> 
  mutate(
    Variable = fct_rev(fct_recode(Variable, !!!playtimeDict)),
    Variable = glue("Playtime ({Variable})")
  ) |> 
  mutate(across(where(is.numeric), ~ round(., 3)))

# write_csv(playtimeTable, "tables-figures/playtimeDescriptives.csv")

@@

Given the lack of holistic or platform-level data available in the literature, our first goal was to simply describe the volume of play. This is visualised in @fig-playtime, which show that despite a total play volume of more than 140k hours, our sample was largely minimally engaged with 1st party Nintendo games. During the two weeks prior to survey completion, just over half of the sample had no sessions logged. The top 10% of players were moderately engaged, playing an average of 60 minutes per day. Sessions of a game lasted on average `r round(median(datRaw$duration, na.rm = T), 1)` minutes \[10th percentile: `r quantile(datRaw$duration, .1)`; 90th percentile: `r quantile(datRaw$duration, .9)`\].

The results of this study are therefore reflective of a largely casual population of players—at least with respect to Nintendo titles. We argue that this population is nonetheless an important one to study alongside the ongoing research focused on so-called hardcore players: if video games were to meaningfully affect well-being, we may expect a larger impact for people who rarely play but happen to play for one hour, than for a highly engaged population of people who tend to play three hours per day, but happen to play four hours. We return to this limitation on generalizability in the discussion.

@@

```{r}
#| label: various-time-models

@@

```{r}
#| label: sensitivity-needsat
#| include: false

datNS <- dat |> 
  mutate(
    as = (bpnsfs_1 + bpnsfs_7 + bpnsfs_13 + bpnsfs_19)/4,
    af = (bpnsfs_2 + bpnsfs_8 + bpnsfs_14 + bpnsfs_20)/4,
    cs = (bpnsfs_3 + bpnsfs_9 + bpnsfs_15 + bpnsfs_21)/4,
    cf = (bpnsfs_4 + bpnsfs_10 + bpnsfs_16 + bpnsfs_22)/4,
    rs = (bpnsfs_5 + bpnsfs_11 + bpnsfs_17 + bpnsfs_23)/4,
    rf = (bpnsfs_6 + bpnsfs_12 + bpnsfs_18 + bpnsfs_24)/4,
    ns = (as + cs + rs)/3,
    nf = (af + cf + rf)/3
  )

NSmodels <- expand.grid(
  predictor = dat |> select(matches("playtime\\d")) |> names(),
  outcome = c("csas", "affectiveValence", "meanWemwbs", "meanPromis")
) |>
  mutate(formula = paste0(outcome, " ~ ", predictor, " + meanLifeFit + ns + nf + age + genderRC + eduLevel + employment"))

fittedNSModels <- map_df(
  NSmodels$formula, ~ lm(as.formula(.x),
    data = datNS
  ) |>
    broom::tidy(conf.int = TRUE, conf.level = .99) |>
    filter(grepl("meanLifeFit", term))
) |>
  mutate(
    predictor = NSmodels$predictor,
    outcome = NSmodels$outcome
  ) |>
  select(predictor, outcome, everything())

nsMod <- lm(ns ~ meanLifeFit + age + genderRC + eduLevel + employment, data = datNS) |> 
  broom::tidy(conf.int = TRUE) |> 
  filter(term == "meanLifeFit") |> 
  mutate(pretty = str_glue("{number(estimate, .01)} 95% CI [{number(conf.low, .01)}, {number(conf.high, .01)}]"))
